Peptide retention time prediction

### 2014-11-30 retention.tex
Small textual changes

Mass spectrometry-based methods are unquestionably the most popular
techniques for studying the proteint content of biological samples.
The palette of protemic applications where such methods are used is
getting increasingly diversified, ranging from identifying and
quantifying proteins in biological samples \cite{Geiger2012} to
determining posttranslational modifications \cite{Huttlin2010} and
elucidating protein interaction networks \cite{Gavin2011}.

@@

Today a majority of the mass spectrometry (MS) protocols in use are
high-throughput, generating extensive amounts of data. This has
brought up a whole new set of analytical challenges, making the
development of efficient algorithms to interpret this data a major
area of proteomics research. A particular difficult challenge in this
context is posed by the tremendous complexity of the data. Proteins
display a wide range of physico-chemical properties and are often
found at very different amounts in complex samples, spanning as much
as 12 orders of magnitude in relative abundance~\cite{Angel2012}.
Furthermore, a large proportion of proteins are subjected to
posttranslational modifications. As an example, a few years ago more
than 25,000 phosphorylation sites on human proteins were already
reported~\cite{Lemeer2009}. To manage this formidable diversity,
increasing efforts are made to extract - besides the fragmentation
spectra - additional information from each MS experiment.

@@

% iRT? \cite{Escher:2012qv}
Following these observations, a variety of retention time prediction
methods has been proposed. Most naturally one could

These methods can be roughly divided into four classes: look-up
tables-based approaches, index-based methods, modeling-based methods
and machine learning techniques. The main ideas underlying each
category are described below, while Table \ref{tab:rtmethods} lists a
few of the most popular retention time prediction methods from each
category.

@@

\section{\label{sec:irt}Peptide retention standards}

\hyphenation{wide-spread}
% peptide standards what they are, what they are used for 
In the past years, the use of retention standards has become
increasingly widespread in the context of mass spectrometry-based
proteomics~\cite{olegstd, irt}. Such standards typically consist of a
small number of peptides covering a wide span of hydrophobicities and
with well studied behaviors under LC-MS/MS. These peptides are often
analyzed together with the sample of interest, and by comparing their
observed and expected behaviors one can obtain valuable information
for a variety of applications.

As an example, in targeted proteomics peptide standards can be used to
 calibrate predicted retention times and to estimate the accuracy of
 the prediction algorithms~\cite{Kiyonami01022011}. In addition,
 recent studies have reported excellent results when using peptide
 standards to adjust the time when certain transitions are scheduled
 ``on-the-fly''~\cite{seb, irt}. Retention standards are also useful
 for aligning data from multiple experiments~\cite{petritis2003}, to
 transfer retention times between different chromatography
 setups~\cite{seb}, or to keep track of the performance of LC-MS/MS
 systems~\cite{qcal}.

@@

Interpretation of mass spectrometry data is a question of
understanding the properties that explain the behavior of peptides in
an mass spectrometry experiment. The available technologies to infer
peptides and proteins from mass spectrometry data involves predicting
properties of peptides in a database, and compare those properties to
the observed data. Such comparisons are for discovery proteomics done
by so called search engines, that traditionally are limited to
predicting predicting mass-to-charge rates of the precursor peptides,
and their fragmentation patterns. However, an attractive way to
increase the sensitivity of the inference procedure is to also predict
properties such as their iso-electric point~\cite{branca2014hirief} or
of particular interest for this review, chromatographic retention
time\cite{cerqueira2010mude}.

The principle of predictability also extends to targeted proteomics,
in SRM analysis the peptides retention time and fragments are
predicted from retention time, often derived from previous shotgun
analysis. As targeted proteomics move into more comprehensive
techniques, such as DIA/SWATH analysis, it becomes increasingly
attractive to use techniques that do not rely on previous observations
of the exact peptide to determine its properties, but instead to use
de-novo predictions from amino acid sequences.

>>>>

Mass spectrometry (MS) based methods are unquestionably the most popular
techniques for studying the protein content of biological samples.
The palette of protemic applications where such methods are used is
getting increasingly diversified, ranging from identifying and
quantifying proteins in biological samples \cite{Geiger2012} to
determining posttranslational modifications \cite{Huttlin2010} and
elucidating protein interaction networks \cite{Gavin2011}.

@@

Today a majority of the mass spectrometry (MS) protocols in use are
high-throughput, generating extensive amounts of data. This has
brought up a whole new set of analytical challenges, making the
development of efficient algorithms to interpret this data a major
area of proteomics research. A particular difficult challenge in this
context is posed by the tremendous complexity of the data. Proteins
display a wide range of physico-chemical properties and are often
found at very different amounts in complex samples, spanning as much
as 12 orders of magnitude in relative abundance~\cite{Angel2012}.
Furthermore, a large proportion of proteins are subjected to
posttranslational modifications. As an example, a few years ago more
than 25,000 phosphorylation sites on human proteins were already
reported~\cite{Lemeer2009}. To manage this formidable diversity,
increasing efforts are made to extract besides the fragmentation
spectra additional information from each MS experiment.

@@

% iRT? \cite{Escher:2012qv}
Following these observations, a variety of retention time prediction
methods has been proposed. Most naturally one could divide these
methods into four classes: look-up tables-based approaches,
index-based methods, modeling-based methods and machine learning
techniques. The main ideas underlying each category are described
below, while Table \ref{tab:rtmethods} lists a few of the most popular
retention time prediction methods from each category.

@@

%\section{\label{sec:irt}Peptide retention standards}

%\hyphenation{wide-spread}
% peptide standards what they are, what they are used for 
%In the past years, the use of retention standards has become
%increasingly widespread in the context of mass spectrometry-based
%proteomics~\cite{olegstd, irt}. Such standards typically consist of a
%small number of peptides covering a wide span of hydrophobicities and
%with well studied behaviors under LC-MS/MS. These peptides are often
%analyzed together with the sample of interest, and by comparing their
%observed and expected behaviors one can obtain valuable information
%for a variety of applications.

%As an example, in targeted proteomics peptide standards can be used to
% calibrate predicted retention times and to estimate the accuracy of
% the prediction algorithms~\cite{Kiyonami01022011}. In addition,
% recent studies have reported excellent results when using peptide
% standards to adjust the time when certain transitions are scheduled
% ``on-the-fly''~\cite{seb, irt}. Retention standards are also useful
% for aligning data from multiple experiments~\cite{petritis2003}, to
% transfer retention times between different chromatography
% setups~\cite{seb}, or to keep track of the performance of LC-MS/MS
% systems~\cite{qcal}.

@@

Interpretation of mass spectrometry data is a question of
understanding the properties that explain the behavior of peptides in
a mass spectrometry experiment. The available technologies to infer
peptides and proteins from mass spectrometry data involve predicting
properties of peptides from a database, and compare those properties to
the observed data. Such comparisons are done for discovery proteomics
by algorithms called search engines, that traditionally are limited to
predicting predicting mass-to-charge rates of the precursor peptides,
and their fragmentation patterns. However, an attractive way to
increase the sensitivity of the inference procedure is to also predict
properties such as their iso-electric point~\cite{branca2014hirief} or
of particular interest for this review, chromatographic retention
time\cite{cerqueira2010mude}.

% This paragraph is consfusing, should be rephrased
The principle of predictability also extends to targeted proteomics,
in SRM analysis the peptides retention time and fragments are
predicted from retention time, often derived from previous shotgun
analysis. As targeted proteomics move into more comprehensive
techniques, such as DIA/SWATH analysis, it becomes increasingly
attractive to use techniques that do not rely on previous observations
of the exact peptide to determine its properties, but instead to use
de-novo predictions from amino acid sequences.
